---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# joyn

<!-- badges: start -->
<!-- badges: end -->

The goal of `joyn` is to provide the user with a set of tools to analyze the quality
of merging (i.e., joining) data frames, so that it is a **JOY** to join tables with
`joyn`. This is inspired in the command `merge` of 
the statistical software `Stata`. Thus, it is not intended to be extremely fast, 
even though it uses the power of `data.table` indexing to perform the merges. The
reason for `joyn` to be a little slower is than pure `data.table` is that it will 
always perform a full join of data in the same way the `Stata` does it. Yet, the 
user can select what data to keep (e.g., that left [master], right [using], 
inner join, or full join). 

`joyn` presents several important features. The most important is the additional
variable, `report`, which summarizes the result of the join (you can change the 
name if you want). In addition, it allows you to specify the kind of join you 
want to perform (i.e, m:m, m:1, 1:m, or 1:1). This ensures that your join does 
not produce unexpected outcome.

## Installation

<!-- You can install the released version of joyn from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("joyn") -->
<!-- ``` -->

The development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("randrescastaneda/joyn")
```
## Example

```{r example}
library(joyn)

x1
y1

# using commong variable `id` as key.
merge(x1, y1)[]

# keep just those observations that match
merge(x1, y1, keep = "inner")[]


x2 

y2


# Bad merge for not specifying by argument
merge(x2, y2)[]

# good merge, ignoring variable x from y
merge(x2, y2, by = "id")[]

# update NAs in x variable form x
merge(x2, y2, by = "id", updateNA = TRUE)[]

# Update values in x with variables from y
merge(x2, y2, by = "id", update_values = TRUE)[]

```

