---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# joyn

<!-- badges: start -->

[![R-CMD-check](https://github.com/randrescastaneda/joyn/workflows/R-CMD-check/badge.svg)](https://github.com/randrescastaneda/joyn/actions)
`r badger::badge_cran_release("joyn", "orange")`
`r badger::badge_devel("randrescastaneda/joyn", "blue")`
`r badger::badge_lifecycle("maturing", "green")`
[![codecov](https://codecov.io/gh/randrescastaneda/joyn/branch/master/graph/badge.svg?token=VTRGFG5H0H)](https://codecov.io/gh/randrescastaneda/joyn)

<!-- badges: end -->

`joyn` provides the user with a set of tools to analyze the quality of merging
(i.e., joining) data frames, so that it is a **JOY** to join tables with `joyn`.
This is inspired on the command `merge` of the statistical software `Stata`.

## Motivation

The objective `joyn` is to make your life easier when joining tables. As a
former Stata user (I still work with Stata but not that much as I work with R
now), I had missed, until now, the ability to assess the accuracy of my join
after mergin two tables in R. With one single command, `merge`, Stata allows the
user to perform any kind of equi-join. The reason for this is that, by default,
Stata merges fully both tables into one and then it is up to the user to keep
the observation she needs. Most importantly, Stata forces the user to know how
her joining tables relate to each other. Most tables have a one-to-one (1:1 )
relation, but it is common to find one-to-many (1:m), many-to-one (m:1), and
many-to-many (m:m) relations. Finally, Stata's `merge` command returns by
default a variable with useful information about the table's join. So, the
following features are the value added of `joyn`:

1.  `joyn` performs a full join by default (i.e., resulting table has all the
    observations from both original, joining tables). Yet, the user can select
    "left", "right" or "inner" joins using the argument `keep`.

2.  In the same vain, `joyn` keeps all the columns form both joining tables. Two
    caveats are worth mentioning.

    1.  If both tables have variables with the same name, by default `joyn` does
        not bring those variables from the second (i.e, right, using) table into
        the first (i.e., left, master) table. Yet, you can do one of two things.
        First, you can use arguments `update_values` or `updateNA` to update the
        value of those variable in the first table with the ones in the second
        table. Or, you can use argument `keep_y_in_x = TRUE` to keep both
        variables with a different name.

    2.  You can use the argument `yvars` to select which variables you want to
        bring from the second table into the first one.

3.  `joyn` allows the user to perform one-to-one (1:1) , one-to-many (1:m),
    many-to-one (m:1), and many-to-many (m:m) joins. The default is m:m,
    following general R's practice, but its is highly discouraged. Since you
    want to use `joyn`, I recommend you always specify the right relationship
    between your joining tables.

4.  `joyn` returns a reporting variable with the status of the join.

## Regarding speed and flexibility

Notice the `joyn` is not intended to be a super fast joining tool. By
construction, it does a lot of things that will make it slower than other tools
out there. However, `joyn` is basically a wrapper around `data.table`'s
`merge.data.table()` function. So, the lost of speed of `joyn` is mainly due to
evaluating several conditions, creating the reporting variable, and present a
nice summary table at the end of the process. Also, keep in mind that `joyn` is
intended to be informative, so it displays messages here and there to inform you
about your join (you can silence any message in `joyn`, including the reporting
table, by using the argument `verbose = FALSE`). This, of course, makes `joyn` a
little slower than using regular `data.table` syntax. However, the loss of speed
is not much and you're gaining a lot of information. The main reason why `joyn`
is a little slower is than pure `data.table` is that it will always perform a
full join of data in the same way `Stata` does. Again, `joyn` is intended for
information and verification of joins of tables. If you are working on a project
that executes many (say, 1000) joins or you are constantly merging super big
data sets, it is recommended that you use `data.table`'s syntaz x directly.

As of now, the flexibility of `joyn` is limited to the basic joins, yet the most
used and useful ones. If you want to learn more about different kinds of joins
available in `data.table` and how they relate to `dplyr`, I recommend you start
with this [blog
post](https://atrebas.github.io/post/2019-03-03-datatable-dplyr/#joinbind-data-sets).

## Installation

You can install the stable version of joyn from
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("joyn")
```

The development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("randrescastaneda/joyn")
```

## Example

```{r example}
library(joyn)

x1
y1

# using commong variable `id` as key.
merge(x1, y1)[]

# keep just those observations that match
merge(x1, y1, keep = "inner")[]


x2 

y2


# Bad merge for not specifying by argument
merge(x2, y2)[]

# good merge, ignoring variable x from y
merge(x2, y2, by = "id")[]

# update NAs in x variable form x
merge(x2, y2, by = "id", update_NAs = TRUE)[]

# Update values in x with variables from y
merge(x2, y2, by = "id", update_values = TRUE)[]


# do not bring any variable from Y into x, just the report
merge(x2, y2, by = "id", yvars = NULL)[]

```
