---
title: "Advanced functionalities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced-functionalities}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)


```

```{r setup}

library(joyn)
library(data.table)

x <- data.table(id = c(1, 4, 2, 3, NA),
                t  = c(1L, 2L, 1L, 2L, NA),
                country = c(16, 12, 3, NA, 15))
  
y <- data.table(id  = c(1, 2, 5, 6, 3),
                gdp = c(11L, 15L, 20L, 13L, 10L),
                country = 16:20)

```

## Advanced use

This vignette will let you explore some additional features available in `joyn`, through an example use case.

Suppose you want to join tables `x` and `y`, where the variable *country* is available in both. You could do one of five things:

### 1. Use variable *country* as one of the key variables

If you don't use the argument `by`, `joyn` will consider *country* and *id* as key variables by default given that they are common between `x` and `y`.

```{r ex1}

# The variables with the same name, `id` and `country`, are used as key
# variables.

joyn(x = x, 
     y = y)

```

Alternatively, you can specify to join by *country*

```{r ex2}

# Joining by country

joyn(x = x, 
     y = y, 
     by = "country")

```

### 2. Ignore the values of *country* from `y` and don't bring it into the resulting table

This the default if you did not include *country* as part of the key variables in argument `by`.

```{r}

joyn(x = x, 
     y = y, 
     by = "id")

```

### 3. Update only NAs in table x

Another possibility is to make use of the `update_NAs` argument of `joyn()`. This allows you to update the NAs values in variable *country* in table `x` with the actual values of the matching observations in *country* from table y. In this case, actual values in *country* from table x will remain unchanged.

```{r ex3}

joyn(x = x,
     y = y, 
     by = "id", 
     update_NAs = TRUE)

```

### 4. Update actual values in table x

You can also update all the values - both NAs and actual - in variable *country* of table `x` with the actual values of the matching observations in *country* from `y`. This is done by setting `update_values = TRUE`.

Notice that the `reportvar` allows you keep track of how the update worked. In this case, *value update* means that only the values that are different between *country* from `x` and *country* from `y` are updated.

However, let's consider other possible cases:

-   If, for the same matching observations, the values between the two *country* variables were the same, the reporting variable would report *x & y* instead (so you know that there is no update to make).

-   if there are NAs in *country* from `y`, the actual values in `x` will be unchanged, and you would see a *not updated* status in the reporting variable. Nevertheless, notice there is another way for you to bring *country* from `y` to `x`. This is done through the argument `keep_y_in_x` (*see 5. below* ⬇️)

```{r ex4}

# Notice that only the value that are 

joyn(x = x, 
     y = y, 
     by = "id", 
     update_values = TRUE)

```

### 5. Keep original *country* variable from y into returning table

#### (Keep matching-names variable from y into x -not updating values in x)

Another available option is that of bringing the original variable *country* from `y` into the resulting table, without using it to update the values in `x`. In order to distinguish *country* from `x` and *country* from `y`, `joyn` will assign a suffix to the variable's name: so that you will get *country.y* and *country.x*. All of this can be done specifying `keep_common_vars = TRUE.`

```{r ex5}

joyn(x = x, 
     y = y, 
     by = "id", 
     keep_common_vars = TRUE)

```

### Bring other variables from y into returning table

In `joyn` , you can also bring non common variables from `y` into the resulting table. In fact you can specify them in `y_vars_to_keep`, as shown in the example below:

```{r ex6}

# Keeping variable gdp 

joyn(x = x, 
     y = y, 
     by = "id", 
     y_vars_to_keep = "gdp")

```

Notice that if you set `y_vars_to_keep = FALSE` or `y_vars_to_keep = NULL`, then `joyn` won't bring any variable into the returning table.
