---
title: "Advanced use"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>", 
  # error    = TRUE, 
  purl     = FALSE
)
```

```{r setup}
library(joyn)
library(data.table)

x <- data.table(
  id      = c(1, 4, 2, 3, NA),
  t       = c(1L, 2L, 1L, 2L, NA),
  country = c(16, 12, 3, NA, 15)
)
  
y <- data.table(
  id      = c(1, 2, 5, 6, 3),
  gdp     = c(11L, 15L, 20L, 13L, 10L),
  country = 16:20
)

```

## Updating variables in master (i.e., left) table

Imagine that variable *country* is available in the two tables that you're going
to join. You could do one of five things:

### 1. Use variable *country* as one of the key variables

This is the default if you don't use the argument `by`

```{r}
# The variables with the same name, `id` and `x`, are used as key
# variables.
merge(x, y)[]
```

### 2. Ignore variable *country* and don't bring it into the resulting table

This the default if you did not include *country* as part of the key variables
in argument `by`.

```{r}
merge(x, y, by = "id")[]
```

### 3. Update only NAs in table x

Update only NAs values in variable *country* in table x (i.e., left or master)
with the actual values of the matching observations in *country* from table y
(i.e, right or using). In this case, actual values in *country* from table x
will remain unchanged.

```{r}
merge(x, y, by = "id", update_NAs = TRUE)[]
```

### 4. Update actual values in table x

Update all the values--NAs and actual--in variable *country* of table x (i.e.,
left or master) with the actual values of the matching observations in *country*
from table y (i.e, right or using). Also, notice the output of the table below:
only the values that are different between *country* from x and *country* from y
are updated. If, for the same matching observations, the values between the two
*country* variables are the same, the reporting variable says "x & y" rather
than "value updated", so you know that there is no update to make. Finally, If
there are NAs in variable *country* from **table y**, the actual values in x
will be **unchanged**, and you get a "not updated" status in the reporting
variable. Yet, there is a way to solve this, which explained below using
argument `keep_y_in_x`.

```{r}
# notice that only the value that are 
merge(x, y, by = "id", update_values = TRUE)[]
```

### 5. Keep matching-names variables from y into x (No update)

Bring variable *country* from y into the resulting table with a different name
(*country.y*).

```{r}
merge(x, y, by = "id", keep_y_in_x = TRUE)[]
```

If you need to update *country* in x, including the actual values with NAs from
*country* in y, you could do the following, keeping in mind that the reporting
variable is not that useful anymore.

```{r}
dd <- merge(x, y, by = "id", keep_y_in_x = TRUE, verbose = FALSE)
dd[, country := country.y
   ][, country.y := NULL]

dd[]
```

## Bring specific variables from y into x

Most of the times, you don't need to join all the variable from table y into
table x, but just a few of them. By default, `joyn` brings all the variables in
y into x because the argument `yvars` is set to `TRUE`. Yet, if you set it to
`FALSE` or `NULL` it will bring NO variable from y into x, only only the
reporting variable, which many times is very nice trick. Finally, if you set the
`yvars` argument to the specific name of variable in y, only that variable will
be carried over in the join.

```{r}
# Bring all variables updating values
merge(x, y, by = "id", update_values = TRUE)[]

# Bring no variables from y
merge(x, y, by = "id", yvars = FALSE)[]

# Bring only variable gdp from table y. So variable country won't be updated either because table y is filtered before the coparison of variables is done. 
merge(x, y, by = "id", yvars = "gdp")[]
```

## Rolling joins

[*Coming soon*]{style="color:red"}

## Misc.

[*Coming soon*]{style="color:red"}
