---
title: "dplyr-joins"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dplyr-joins}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

withr::local_options(joyn.verbose = TRUE)

```

## Overview

Joining data tables with `joyn` is particularly convenient as it allows you to analyze/be aware of the quality of the merging.

This vignette explores dplyr-like join functions available in `joyn`. Their major objective is to let you employ a syntax you are supposedly already familiar with - the `dplyr` one - while at the same time benefiting of the additional tools that `joyn` offers. That is, obtaining additional information and verification of the joining.

There are four types of dplyr-like join functions

-   Left joins: `joyn::left_join()`

-   Right joins: `joyn::right_join()`

-   Full joins: `joyn::full_join()`

-   Inner joins: `joyn::inner_join()`

Each of them is a wrapper that works in a similar way as the corresponding `dplyr` function.

```{r setup}

library(joyn)
library(data.table)

```

## Rationale

```{r}

x1 <- data.table(id  = c(1L, 1L, 2L, 3L, NA_integer_),
                 t   = c(1L, 2L, 1L, 2L, NA_integer_),
                 x   = 11:15)

y1 <- data.table(id  = c(1,2, 4),
                 y   = c(11L, 15L, 16))


```

Suppose you want to perform a **simple left join** between tables `x1` and `y1`.

With `joyn` you have two possibilities:

-   using the `joyn()` function, specifying `keep = "left"`

-   using the **`joyn::left_join()`** function

In addition, you could use `dplyr::left_join()` or base R merging functions.

Consider these three options:

```{r}

# Option 1

joyn(x          = x1, 
     y          = y1, 
     keep       = "left",
     match_type = "m:1")

# Option 2 

joyn::left_join(x            = x1, 
                y            = y1, 
                relationship = "many-to-one") 

# Option 3

dplyr::left_join(x            = x1, 
                 y            = y1, 
                 relationship = "many-to-one") 


```

As a result, the same returning data table will be produced.

However, using `joyn::left_join()` allows you to enjoy the benefits of the other two options: the intuitive syntax from `dplyr` and the additional tools from `joyn`. These include additional arguments to be used, the availability of the joyn report, messages informing you on time of execution and the status of the join as well as the execution of various checks during the merging.

## Use cases 

#### Left join:

Left join will return all rows from the left table and the matching rows from the right table.

```{r left-ex1}

# Data tables to be joined 
df1 <- data.frame(id = c(1L, 1L, 2L, 3L, NA_integer_, NA_integer_),
                  t  = c(1L, 2L, 1L, 2L, NA_integer_, 4L),
                  x  = 11:16)

df2 <- data.frame(id = c(1,2, 4, NA_integer_, 8),
                  y  = c(11L, 15L, 16, 17L, 18L))

left_join(x = df1, 
          y = df2,
          relationship = "many-to-one",
          update_NAs = TRUE)


```

#### Right join

#### Full join 

#### Inner join 
