---
title: "General use"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{General use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>", 
  error    = TRUE, 
  purl     = FALSE
)
```

# Motivation

The objective `joyn` is to make your life easier when joining tables. As a
former Stata user (I still work with Stata but not that much as I work with R
now), I have missed the ability to understand the results of joining tables.
With one single command, `merge`, Stata allows the user to perform any kind of
equi-join. The reason for this is that, by default, Stata merges fully both
tables into one and then it is up to the user to keep the observation she needs.
Most importantly, Stata forces the user to know how her tables relate to each
other. Some tables have a one-to-one (1:1 ) relation, but it is common to find
one-to-many (1:m), many-to-one (m:1), and many-to-many (m:m) relations. Finally,
Stata's `merge` command returns by default a variable with useful information
about the table's join. So the following features are the value added of `joyn`:

1.  Perform a full join by default (i.e., final table has all the observations
    of both original tables). Yet, the user can select "left", "right" or
    "inner" joins using the argument `keep`.

2.  In the same vain, `joyn` merges all the columns in both tables. To caveats
    are worth mentioning.

    1.  If both tables have variables with the same name, by default `joyn` does
        not bring those variables from the second (i.e, right, using) table into
        the first (i.e., left, master) table. Yet, you can do one of two things.
        First, you can use arguments `update_values` or `updateNA` to update the
        value of those variable in the first table with the ones in the second
        table. Or, you can use argument `keep_y_in_x = TRUE` to keep both
        variables with a different name.

    2.  You can use the argument `yvars` to select which variables you want to
        bring from the second table into the first one.

3.  Allow the user to perform one-to-one (1:1) , one-to-many (1:m), many-to-one
    (m:1), and many-to-many (m:m) joins. The default is m:m, following general
    R's practice, but its use is highly discouraged.

4.  Return a reporting variable with the status of the join.

# Regarding speed and flexibility

Notice the `joyn` is not intended to be a super fast joining tool. By
construction it does a lot of things that will make it slower that other tools
out there. However, `joyn` is basically a wrapper around `data.table`'s indexed
joining tools. So, the lost of speed of `joyn` is mainly due to asses some
conditions, create the reporting variable, and (what really takes the most time)
present a nice summary table when `joyn` concludes. Also, keep in mind that
`joyn` is intended to be informative, so it displays messages here and there to
inform you about your join (you can silence any message in `joyn`, including the
reporting table, by using the argument `verbose = FALSE`).

As of now, the flexibility of `joyn` is limited to the basic joins, yet the most
used and useful ones. If you want to learn more about different kinds of joins
available in `data.table` and how they relate to `dplyr`, I recommend you start
with this [blog
post](https://atrebas.github.io/post/2019-03-03-datatable-dplyr/#joinbind-data-sets).


```{r setup}
library(joyn)
library(data.table)
```
