---
title: "Messages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Messages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

#withr::local_options(joyn.verbose = FALSE)
#withr::local_options(joyn.match_type = "1:1")
withr::local_options(joyn.verbose = TRUE)

```

✅ This vignette is dedicated to one specific set of additional tools that joyn provides you : **messages**.

Messages in joyn are intended to inform you about the status of your join, so that you can better understand what the merge does, ... XXXX TODO.

We'll start with a rough overview of the different kinds of messages you can get when merging two data tables, then discuss each of them in detail with representative examples.

# Overview

Joyn messages can be of 4 different types:

1.  <span style = "color: #228B22;">**Info**

2.  `<span style = "color: #`{=html}0000B8`;">`{=html}**Timing**

3.  `<span style = "color: #`{=html}FE5A1D`;">`{=html}**Warning**

4.  `<span style = "color: #`{=html}CE2029`;">`{=html}**Error**

```{r msg-type, eval = TRUE}

# Setup
library(joyn)

# Checking available types of messages
msgs_types = joyn:::type_choices()

```
