---
title: "Messages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Messages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

#withr::local_options(joyn.verbose = FALSE)
#withr::local_options(joyn.match_type = "1:1")
withr::local_options(joyn.verbose = TRUE)

```

✅ This vignette is dedicated to one specific feature of joyn: displaying information through **messages**.

Messages in joyn are intended to inform you about the status of your join, so that you can better understand what the merge does, ... XXXX TODO.

We'll start with a rough overview of the different kinds of messages that might be generated when merging two data tables, then discuss each of them in detail with representative examples.

## Overview

Joyn messages can be of 4 different types:

1.  <span style = "color: #228B22;">**Info**

2.  `<span style = "color: #`{=html}0000B8`;">`{=html}**Timing**

3.  `<span style = "color: #`{=html}FE5A1D`;">`{=html}**Warning**

4.  `<span style = "color: #`{=html}CE2029`;">`{=html}**Error**

```{r setup}

# Setup 
library(joyn)
library(data.table)
library(cli)

```


```{r msg-type}

# Checking available types of messages
msgs_types = joyn:::type_choices()
print(msgs_types)

```

### Information messages

Info messages inform you about XXX TODO, as you can see in the examples below. 

Recall that one of the additional features of joyn is that it returns a reporting variable with the status of the join. Info messages are generated to tell you in which variable it is available the joyn report, or if the reporting variable is not returned instead. 

```{r info-ex1, eval = FALSE}

# Example dataframes

x1 = data.table(id = c(1L, 1L, 2L, 3L, NA_integer_),
                t  = c(1L, 2L, 1L, 2L, NA_integer_),
                x  = 11:15)

y1 = data.table(id = c(1,2, 4),
                y  = c(11L, 15L, 16))


x2 = data.table(id = c(1, 4, 2, 3, NA),
                t  = c(1L, 2L, 1L, 2L, NA_integer_),
                x  = c(16, 12, NA, NA, 15))


y2 = data.table(id = c(1, 2, 5, 6, 3),
                yd = c(1, 2, 5, 6, 3),
                y  = c(11L, 15L, 20L, 13L, 10L),
                x  = c(16:20))


# ------------------- Showing which var contains joyn report -------------------

# Joining x1 and y1
joyn(x = x2,
     y = y2,
     by = "id", 
     y_vars_to_keep = FALSE)

# Printing the info message
print(joyn_msg(type = "info"))

# ---------------- Info about change in reporting variable name ----------------

```

Also, an info message might be generated to let you know that the name you have specified (?you want to assign) for the reporting variable is already present in the returning table so that it will be changed to a valid one.

```{r info-ex2}

```


### Timing messages

### Warning messages

### Error messages

