---
title: "Performance benchmarks"
description: >
  How fast is joyn compared to dplyr, data.table, collapse, and base R?
  This article presents benchmarks across table sizes and explains the 
  performance tradeoffs of joyn's diagnostic layer.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE
)

read_benchmark_artifact <- function() {
  candidates <- c(
    file.path("inst", "benchmarks", "results", "benchmark-results-latest.rds"),
    file.path("..", "inst", "benchmarks", "results", "benchmark-results-latest.rds"),
    file.path("..", "..", "inst", "benchmarks", "results", "benchmark-results-latest.rds"),
    system.file("benchmarks", "results", "benchmark-results-latest.rds", package = "joyn")
  )

  candidates <- unique(candidates[nzchar(candidates)])
  existing <- candidates[file.exists(candidates)]

  if (length(existing) == 0) {
    stop(
      paste(
        "Benchmark artifact not found.",
        "Run `Rscript inst/benchmarks/bench_joyn_microbenchmark.R` from the package root first.",
        sep = "\n"
      ),
      call. = FALSE
    )
  }

  readRDS(existing[[1]])
}

format_n_label <- function(x) {
  ifelse(
    x >= 1e6,
    paste0(format(x / 1e6, trim = TRUE), "M"),
    paste0(format(x / 1e3, trim = TRUE), "K")
  )
}

benchmark_artifact <- read_benchmark_artifact()
benchmark_results <- benchmark_artifact$results
benchmark_measurements <- benchmark_artifact$measurements
benchmark_meta <- benchmark_artifact$metadata

benchmark_results$n_label <- paste0("n = ", format_n_label(benchmark_results$n))
benchmark_measurements$n_label <- paste0("n = ", format_n_label(benchmark_measurements$n))

n_levels <- paste0("n = ", format_n_label(benchmark_meta$sizes))
benchmark_results$n_label <- factor(benchmark_results$n_label, levels = n_levels)
benchmark_measurements$n_label <- factor(benchmark_measurements$n_label, levels = n_levels)

join_levels <- c("left join", "right join", "full join", "inner join", "anti join")
benchmark_results$join_label <- factor(benchmark_results$join_label, levels = join_levels)
benchmark_measurements$join_label <- factor(benchmark_measurements$join_label, levels = join_levels)

tool_levels <- c(
  "collapse::join",
  "dplyr::left_join",
  "joyn::left_join",
  "data.table merge method",
  "base::merge",
  "dplyr::right_join",
  "joyn::right_join",
  "dplyr::full_join",
  "joyn::full_join",
  "dplyr::inner_join",
  "joyn::inner_join",
  "dplyr::anti_join",
  "joyn::anti_join",
  "data.table anti join",
  "base anti join"
)
benchmark_results$tool <- factor(benchmark_results$tool, levels = tool_levels)
benchmark_measurements$tool <- factor(benchmark_measurements$tool, levels = tool_levels)

palette_tools <- c(
  "collapse::join" = "#355070",
  "dplyr::left_join" = "#6D597A",
  "joyn::left_join" = "#B56576",
  "data.table merge method" = "#E09F3E",
  "base::merge" = "#81B29A",
  "dplyr::right_join" = "#6D597A",
  "joyn::right_join" = "#B56576",
  "dplyr::full_join" = "#6D597A",
  "joyn::full_join" = "#B56576",
  "dplyr::inner_join" = "#6D597A",
  "joyn::inner_join" = "#B56576",
  "dplyr::anti_join" = "#6D597A",
  "joyn::anti_join" = "#B56576",
  "data.table anti join" = "#E09F3E",
  "base anti join" = "#81B29A"
)

build_boxplot_chart <- function(data, title, subtitle = NULL) {
  chart_data <- data
  tool_order <- unique(as.character(chart_data$tool))
  chart_data$expr <- factor(chart_data$tool, levels = tool_order)

  hc_dt <- highcharter::data_to_boxplot(
    chart_data, time, expr,
    add_outliers = FALSE,
    name = "Time in milliseconds"
  )

  series_colors <- unname(palette_tools[tool_order])
  for (i in seq_along(hc_dt$data[[1]])) {
    hc_dt$data[[1]][[i]]$color <- series_colors[[i]]
  }

  chart <- highcharter::highchart() |>
    highcharter::hc_xAxis(type = "category") |>
    highcharter::hc_chart(inverted = TRUE) |>
    highcharter::hc_title(text = title) |>
    highcharter::hc_yAxis(title = list(text = "Time (ms)"), min = 0) |>
    highcharter::hc_tooltip(
      shared = FALSE,
      pointFormat = paste0(
        "<b>{point.name}</b><br/>",
        "Low: {point.low:.2f}<br/>",
        "Q1: {point.q1:.2f}<br/>",
        "Median: {point.median:.2f}<br/>",
        "Q3: {point.q3:.2f}<br/>",
        "High: {point.high:.2f}"
      )
    ) |>
    highcharter::hc_legend(enabled = FALSE) |>
    highcharter::hc_add_series_list(hc_dt)

  if (!is.null(subtitle)) {
    chart <- chart |> highcharter::hc_subtitle(text = subtitle)
  }

  chart
}
```

## Overview

`joyn` is built on top of [`collapse`](https://sebkrantz.github.io/collapse/) and
[`data.table`](https://rdatatable.gitlab.io/data.table/) for its join engine, and
adds a **diagnostic layer**: match-type validation, reporting (the `.joyn` variable),
and structured message logging. This layer is what sets `joyn` apart from a plain
merge — but it also comes at a cost.

This article presents two complementary benchmark views:

- **Cross-engine comparison** for the join type all tools share most naturally: a
  left join.
- **Complete join-type coverage** for left, right, full, inner, and anti joins
  using the latest benchmark artifact generated from `inst/benchmarks/bench_joyn_microbenchmark.R`.

The numbers below are not hard-coded. The vignette reads the current benchmark
artifact directly, so rerunning the benchmark script refreshes the tables and plots.

> **Bottom line:** `joyn` is comparable in speed to `data.table` joins and
> `base::merge()` in the common left-join case. Relative to raw `collapse::join`,
> the extra time is the price of diagnostics — match-type checking, `.joyn`
> report generation, and message logging. That overhead is small and mostly fixed,
> so it matters less as table size grows.

---

## Benchmark setup

All benchmarks were run with:

- **Platform:** `r benchmark_meta$platform`
- **R version:** `r benchmark_meta$r_version`
- **Packages:** `r paste(names(benchmark_meta$package_versions), benchmark_meta$package_versions, collapse = ", ")`
- **Key structure:** single integer key (`id`), partially overlapping 1:1 key space
- **Shared-key overlap:** `r sprintf("%.0f%%", 100 * benchmark_meta$overlap)`
- **Sizes:** `r paste(format(benchmark_meta$sizes, big.mark = ","), collapse = ", ")`
- **Repetitions:** small `r benchmark_meta$repetitions$small`, medium `r benchmark_meta$repetitions$medium`, large `r benchmark_meta$repetitions$large`, extra-large `r benchmark_meta$repetitions$xl`
- **Metric:** median elapsed time (ms), with full sampled distributions retained for plotting
- **Generated:** `r benchmark_meta$generated_at`

Data are generated as two random-keyed tables with unique integer keys in each
input and controlled overlap between `x` and `y`. That makes left, right, full,
inner, and anti joins all meaningful in the same benchmark run.

For left, right, full, and inner joins, the comparison includes `joyn`, `dplyr`,
`collapse`, explicit `data.table merge method()`, and base R. For anti joins,
`data.table` and base R are benchmarked using equivalent filtering expressions.

## Results

### Cross-engine comparison: left join

```{r results-table}
library(knitr)
library(data.table)

left_results <- benchmark_results[benchmark_results$join_type == "left", , drop = FALSE]
left_table <- data.table::dcast(
  data.table::as.data.table(left_results[, c("tool", "n_label", "median")]),
  tool ~ n_label,
  value.var = "median"
)
left_table <- as.data.frame(left_table)
left_table <- left_table[match(
  c(
    "collapse::join",
    "dplyr::left_join",
    "joyn::left_join",
    "data.table merge method",
    "base::merge"
  ),
  left_table$tool
), , drop = FALSE]
names(left_table)[1] <- "Tool"

kable(
  left_table,
  digits = 1,
  align = c("l", rep("r", ncol(left_table) - 1L)),
  caption = "Median elapsed time (ms) — left join on a single integer key"
)
```

As expected, `collapse::join()` is the raw-speed leader. `joyn::left_join()` is
slower because it performs validation and reporting work around the underlying
join, but it stays in the same broad range as explicit `data.table merge method()`
and `base::merge()` implementations.

### Complete join-type coverage

```{r complete-results-table}
complete_table <- data.table::dcast(
  data.table::as.data.table(benchmark_results[, c("join_label", "tool", "n_label", "median")]),
  join_label + tool ~ n_label,
  value.var = "median"
)
complete_table <- as.data.frame(complete_table)
names(complete_table)[1:2] <- c("Join", "Tool")

kable(
  complete_table,
  digits = 1,
  align = c("l", "l", rep("r", ncol(complete_table) - 2L)),
  caption = "Median elapsed time (ms) — all benchmarked join types and engines"
)
```

This table is generated from the same latest artifact used for the plots, so it
updates automatically whenever the benchmark runner is re-executed.

### joyn wrapper summary

```{r wrapper-results-table}
joyn_table <- data.table::dcast(
  data.table::as.data.table(
    benchmark_results[
      benchmark_results$engine == "joyn",
      c("join_label", "n_label", "median")
    ]
  ),
  join_label ~ n_label,
  value.var = "median"
)
joyn_table <- as.data.frame(joyn_table)
names(joyn_table)[1] <- "Join"

kable(
  joyn_table,
  digits = 1,
  align = c("l", rep("r", ncol(joyn_table) - 1L)),
  caption = "Median elapsed time (ms) — joyn wrapper joins"
)
```

---

## Distribution view

The benchmark runner now retains the full set of sampled timings, so the charts
below show the spread of elapsed times rather than only the medians.

```{r left-join-chart, fig.width = 8.5, fig.height = 4.8, fig.alt = "Interactive boxplot of left-join benchmark times across engines."}
left_measurements <- benchmark_measurements[
  benchmark_measurements$join_type == "left" &
    benchmark_measurements$n == max(benchmark_measurements$n),
  ,
  drop = FALSE
]
left_measurements <- left_measurements[order(left_measurements$tool), , drop = FALSE]

build_boxplot_chart(
  data = left_measurements,
  title = paste0("Left join distribution at ", unique(as.character(left_measurements$n_label))),
  subtitle = "Each box summarizes the sampled microbenchmark timings for one engine."
)
```

```{r all-joins-chart, fig.width = 10, fig.height = 9, fig.alt = "Interactive boxplots for benchmark times by join type."}
chart_list <- lapply(levels(benchmark_measurements$join_label), function(join_name) {
  chart_data <- benchmark_measurements[
    benchmark_measurements$join_label == join_name &
      benchmark_measurements$n == max(benchmark_measurements$n),
    ,
    drop = FALSE
  ]

  if (nrow(chart_data) == 0) {
    return(NULL)
  }

  build_boxplot_chart(
    data = chart_data,
    title = join_name,
    subtitle = paste0("Distribution at ", unique(as.character(chart_data$n_label)))
  )
})

htmltools::tagList(Filter(Negate(is.null), chart_list))
```

---

## What drives joyn's overhead?

joyn does more than a plain merge on every call:

1. **Input validation** — checks that both tables are non-empty, that `by` variables
   exist and have supported classes.
2. **Match-type enforcement** — verifies the stated `match_type` (e.g., `"1:1"`)
   against the actual data, alerting you if duplicates are present.
3. **Report generation** — builds the `.joyn` factor column that tracks which rows
   came from `x`, `y`, or both.
4. **Message logging** — stores structured info/warning/error messages for later
   retrieval via `joyn_msg()`.

These steps add a modest fixed overhead per call, which is the direct cost of
getting join diagnostics automatically. As table size increases, this overhead
becomes a smaller fraction of total time.

---

## When to use each tool

| Tool | Use when… |
|---|---|
| **joyn** | You want join diagnostics, match-type validation, and a report — especially when data quality matters |
| **collapse::join** | You need maximum raw speed and direct support for left, right, full, inner, and anti joins |
| **dplyr** | You are already in a tidyverse pipeline and do not need diagnostics |
| **data.table** | You are working in data.table and want low-level control, including explicit joins and anti-join idioms |
| **base** | You want dependency-free joins or simple filtering-based anti-join logic |

> **Tip:** `joyn` and `collapse` are complementary. Use `joyn` during data
> preparation and validation; switch to `collapse::join` directly in
> production pipelines where correctness has already been established.

---

## Reproducing these benchmarks

The benchmark artifact used in this vignette is generated by
`inst/benchmarks/bench_joyn_microbenchmark.R`.

```r
# From the package root, in a terminal
Rscript inst/benchmarks/bench_joyn_microbenchmark.R
```

The latest artifact is written to:

- `inst/benchmarks/results/benchmark-results-latest.rds`
- `inst/benchmarks/results/benchmark-results-latest.csv`
- `inst/benchmarks/results/benchmark-measurements-latest.csv`
